---
title: "MotifMatchr"
format: html
editor: visual
---

From: https://github.com/GreenleafLab/motifmatchr and https://greenleaflab.github.io/motifmatchr/articles/motifmatchr.html#motifmatchr

# Install Motifmatchr

```{r}

#BiocManager::install(c("TFBSTools", "motifmatchr", "BSgenome.Mmusculus.UCSC.mm10", "GenomicRanges"))
#BiocManager::install(c("SummarizedExperiment"))
#BiocManager::install("universalmotif")
#BiocManager::install("BSgenome.Rnorvegicus.UCSC.rn6")

library(motifmatchr)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Rnorvegicus.UCSC.rn6)
library(GenomicRanges)
library(SummarizedExperiment)
library(universalmotif)
library(BiocParallel)

```

# Load data

Mandatory inputs: 
1. Position weight matrices (PWM) stored as an object from the TFBSTools package 
2. Either a set of genomic ranges or a set of sequences

```{r}

# Genomic Ranges (peaks) in GRanges object format 
#consensus_chromatin_landscape <- rtracklayer::import("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/consensus_chromatin_landscape.bed")
#consensus_chromatin_landscape <- rtracklayer::import("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/exp24_consensus_peaks_sorted_clean.bed")

# Prolactin specific - mm10 reduced regions
consensus_chromatin_landscape <- rtracklayer::import("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/consensus_chromatin_landscape_chr13.bed")
# Prolactin specific - rn6 reduced regions
#consensus_chromatin_landscape <- rtracklayer::import("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/exp24_consensus_peaks_sorted_clean_chr17.bed")

# Genome
genome(consensus_chromatin_landscape) <- "mm10"
#genome(consensus_chromatin_landscape) <- "rn6"

# Motif sequences
jaspar_2026="/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/JASPAR_CORE_2026_non-redundant.meme"
motifs <- read_meme(jaspar_2026)
  
# convert to PFM for motifmatchr input
pfm_individual <- convert_motifs(motifs, class = "TFBSTools-PFMatrix")
pfm_motifs <- do.call(PFMatrixList, pfm_individual)
class(pfm_motifs)
# ID and names are swapped, fix:
pfm_list_swapped <- lapply(pfm_motifs, function(motif_name) {
  old_id <- motif_name@ID
  old_name <- motif_name@name
  motif_name@ID <- old_name
  motif_name@name <- old_id
  return(motif_name)
})
pfm_motifs <- do.call(PFMatrixList, pfm_list_swapped)
# Add names to the motifs
names(pfm_motifs) <- sapply(pfm_motifs, function(x) paste0(x@ID, "_", x@name))

```

# Inputs

```{r}

motifs_object <- pfm_motifs
peaks <- consensus_chromatin_landscape

```

# Get motif matches in peaks

```{r}

register(MulticoreParam(4))

motif_ix <- matchMotifs(motifs_object,
                        peaks,
                        genome = BSgenome.Mmusculus.UCSC.mm10,
                        p.cutoff = 5e-2)
#motifMatches(motif_ix)

# Save object
saveRDS(motif_ix, file = "/Users/georgiagoddard/Desktop/rn6_prl_motifs/motif_matches_prl_mm10_p0.05.rds")
```

# Get motif poisitions within peaks

```{r}

register(MulticoreParam(8))

motif_pos <- matchMotifs(motifs_object,
                         peaks,
                         genome = BSgenome.Mmusculus.UCSC.mm10,
                         p.cutoff = 5e-2,
                         out = "positions")

# Save object
saveRDS(motif_pos, file = "/Users/georgiagoddard/Desktop/mm10_prl_motifs/motif_positions_prl_mm10_p0.05.rds")

```

```{r}
# The following code is for the color scale bit
all_matches <- unlist(motif_pos)
all_matches$score <- all_matches$score 

rtracklayer::export(all_matches, "/Users/georgiagoddard/Desktop/mm10_prl_motifs/motif_matches.bed", format = "BED")


```

# Read Objects

```{r}

# Read objects in
motif_ix <- readRDS("/Users/georgiagoddard/Desktop/mm10_prl_motifs/motif_matches_prl_mm10_p0.05.rds")
#motif_ix_mm10 <- readRDS("/Users/georgiagoddard/Downloads/all_motif_matches_mm10.rds")
motif_pos <- readRDS("/Users/georgiagoddard/Desktop/mm10_prl_motifs/motif_positions_prl_mm10_p0.05.rds")

```

# Download Outputs

```{r}
# Matches
match_df <- as.data.frame(as.matrix(motifMatches(motif_ix)))
rownames(match_df) <- paste0(seqnames(motif_ix), ":", start(motif_ix), "-", end(motif_ix))
write.csv(match_df, "/Users/georgiagoddard/Desktop/rn6_prl_motifs/Peak_motif_matches.csv")

# Positions
motif_bed <- unlist(motif_pos)
mcols(motif_bed)$name <- names(motif_bed)
export.bed(motif_bed, "/Users/georgiagoddard/Desktop/mm10_prl_motifs/Peak_motif_positions.bed")

# Create a subfolder to keep your Desktop clean
output_dir <- "/Users/georgiagoddard/Desktop/mm10_prl_motifs/splits_0.05_scores"

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

for (motif_id in names(motif_pos)) {

  current_motif <- motif_pos[[motif_id]]
  if (length(current_motif) > 0) {
  
    seqlevelsStyle(current_motif) <- "UCSC"
    clean_name <- gsub("[^[:alnum:]]", "_", motif_id)
    file_path <- file.path(output_dir, paste0(clean_name, ".bed"))
    #export.bed(current_motif, con = file_path)
    rtracklayer::export(current_motif, file_path, format = "BED")
  }
}

message("Done.")

```

```{bash}
# In terminal
# cd /Users/georgiagoddard/Desktop/Research_Project/TF_mapping/data/Motifs/
# Sort and compress (if not already done)
#sort -k1,1 -k2,2n Peak_motif_positions.bed > Peak_motif_positions_sorted.bed
#bgzip Peak_motif_positions_sorted.bed
# Create the index (.tbi)
#tabix -p bed Peak_motif_positions_sorted.bed.gz
#rm Peak_motif_positions.bed
```

# Load Positions back in

```{r}

# Import positions file
motif_bed_imported <- import("/Users/georgiagoddard/Desktop/mm10_prl_motifs/Peak_motif_positions_sorted.bed.gz")
```

```{r}
# Build GRanges object 
motif_pos <- split(motif_bed_imported, mcols(motif_bed_imported)$name)
print(length(motif_pos))
print(head(names(motif_pos)))
```

## Download one BED file per motif

```{r}

# Create a subfolder to keep your Desktop clean
output_dir <- "/Users/georgiagoddard/Desktop/mm10_prl_motifs/splits_0.05_scores"

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

for (motif_id in names(motif_pos)) {

  current_motif <- motif_pos[[motif_id]]
  if (length(current_motif) > 0) {
  
    seqlevelsStyle(current_motif) <- "UCSC"
    clean_name <- gsub("[^[:alnum:]]", "_", motif_id)
    file_path <- file.path(output_dir, paste0(clean_name, ".bed"))
    export.bed(current_motif, con = file_path)
  }
}

message("Done.")

```

# Look for Motifs in target peaks

```{r}

# Load match_df matrix into R studio

# Using the match_df file
# We can select for specific chromosomes or maybe even a range of values/positions
# Or select for a motif name (?) and "TRUE" to see where it's present


```
