---
title: "TF_expression_PCA"
format: html
editor: visual
---

**This script filters the JASPAR 2026 list of 2059 TF motifs to 733 biologically relevant TF motifs**

### Introduction

The marginal footprints command takes too long to load 2059 TFs per cell type per cohort. A lot of these TF motifs will be irrelevant for the pituitary cell types anyway.

The CPA was queried to extract `median_RNAexpression_per_celltype.csv` which contains median gene expression across every gene in each anterior pituitary cell type used throughout this project.

#### Pipeline Breakdown

Within the following script, I identify:

-   The JAPSAR 2026 non-redundant database has 2059 TF motifs that cover 952 TFs (some TFs have multiple motifs).
-   Within the median expression data, only 707 of the 952 TFs have RNA expression data.
-   The 245 TFs with no RNA expression data are a majority zinc finger proteins.
-   The 707 TFs with gene expression data are reduced to a list of 436 using a threshold of minimum median RNA expression \>0.4 in at least 1 cell type.
-   The 245 missing TFs are added back into this list, to make a total of 681 TFs.
-   These TFs expand to 733 total TF motifs taking into account the one to many relationship of some TFs.

### HPC Parallelisation

This list of 733 TF motifs was split into 3 separate lists to be run in parallel using the `MarginalFootprints` command (`chrombpnet_pipeline`) on the HPC cluster, for the sake of speed.

### Libraries

```{r}
library(universalmotif)
library(motifmatchr)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10)
library(GenomicRanges)
library(Matrix)
library(uwot)
library(ggplot2)
library(rhdf5)
```

### RNA-seq gene expression

#### Load RNA-seq median expression file

```{r}

# Load files
RNA_exp_celltype <- read.csv("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/rawdata/median_RNAexpression_per_celltype.csv")
head(RNA_exp_celltype)
JASPAR_tf_motifs <- read_meme("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/JASPAR_CORE_2026_non-redundant.meme")

# Assign rownames to gene ID
rownames(RNA_exp_celltype) <- RNA_exp_celltype$X
RNA_exp_celltype$X <- NULL

# See overlapping TF names
TF_alt_names <- sapply(JASPAR_tf_motifs, function(x) x@altname)
TF_alt_names_lowered <- tolower(TF_alt_names)
matching_rows <- rownames(RNA_exp_celltype)[tolower(rownames(RNA_exp_celltype)) %in% TF_alt_names_lowered]
missing_rows <- setdiff(TF_alt_names_lowered, tolower(rownames(RNA_exp_celltype)))
znf_rows <- grep("znf", missing_rows, value = TRUE)
combo_motif_rows <- grep("::", missing_rows, value = TRUE)
rest_of_missing_motifs <- setdiff(missing_rows, c(znf_rows, combo_motif_rows))

# Print results
cat("There are ", length(TF_alt_names_lowered), "in the JASPAR vertebrate non-redundant TF list.\nBut this includes multiple motifs per TF if there are multiple.\n")
cat("There are only really", length(unique(TF_alt_names_lowered)), "TFs in this list.\n")
cat("Found", length(matching_rows), "matching TFs.\n\n")
cat("There are", length(missing_rows), "missing TFs from the RNA-seq expression list.\n")
# Only 952 TFs really...
# Missing rows are zinc fingers, and dimer motifs. To be kept in downstream analysis. 
cat("There are", length(znf_rows), "zinc finger proteins within this list of missing TFs.\n")
cat("There are", length(combo_motif_rows), "motif combination TFs in this list of missing TFs.\n")
cat("The rest of the missing motifs are:", rest_of_missing_motifs, "\n")
cat("All missing TF motifs will still be incldued in the footprinting.")

```

#### Filter for 707 matching TFs

```{r}
RNA_exp_celltype_TF <- RNA_exp_celltype[matching_rows, , drop = FALSE]

# Keep rows that have a median expression >0.4 in at least one cell type (Prop1 only has max 0.6)
keep_rows <- rowSums(RNA_exp_celltype_TF > 0.4) > 0
RNA_exp_celltype_TF_threshold <- RNA_exp_celltype_TF[keep_rows, , drop = FALSE]
# Threshold 0.0 returns 519 TFs -> 188 TFs with 0 expression in any cell type
# Threshold 0.4 returns 436 TFs 
# Threshold 0.5 returns 425 TFs

```

436 expressed TFs + 245 TFs in JASPAR not included in gene expression list -\> 681 total TFs

But how many does this become if we take into account the fact that some TFs have multiple motifs?

### Make list of final TFs to plot:

```{r}

# List of expressed TFs 
expressed_tfs_lowered <- tolower(rownames(RNA_exp_celltype_TF_threshold))
# Missing TFs
missing_tfs_lowered <- unique(missing_rows)
# Combine for FINAL LIST
final_tfs_lowered <- c(expressed_tfs_lowered, missing_tfs_lowered)

# Filter JASPAR to find motifs associated with TF
final_motifs_list <- JASPAR_tf_motifs[TF_alt_names_lowered %in% final_tfs_lowered]
final_motif_ids <- sapply(final_motifs_list, function(x) x@name)
final_motif_names <- sapply(final_motifs_list, function(x) x@altname)

final_motifs_manifest <- data.frame(
  Motif_ID = final_motif_ids,
  TF_Name = final_motif_names,
  stringsAsFactors = FALSE
)

cat("The 681 unique TFs expand out to a total of", length(final_motifs_list), "motifs in the JASPAR file.\n")
head(final_motifs_manifest)
```

That's too many to plot -\> so we will pick the consensus motifs from JASPAR redundant list instead

```{r}

JASPAR_tf_motifs_redundant <- read_meme("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/JASPAR_CORE_2026_redundant.meme")

redundant_tf_names <- sapply(JASPAR_tf_motifs_redundant, function(x) x@altname)
redundant_ma_ids <- sapply(JASPAR_tf_motifs_redundant, function(x) x@name)

# Lowercase
redundant_tf_names_lowered <- tolower(redundant_tf_names)

# 4. Use your existing 681 unique lowercase TF list from the previous steps
# (Combining the 436 expressed and 245 missing TFs)
expressed_tfs_lowered <- tolower(rownames(RNA_exp_celltype_TF_threshold))
missing_tfs_lowered <- unique(missing_rows)
final_tfs_lowered <- c(expressed_tfs_lowered, missing_tfs_lowered)

# 5. Filter the redundant keys based on your target TF list
matching_indices <- which(redundant_tf_names_lowered %in% final_tfs_lowered)

# 6. Create the final clean mapping table
final_redundant_manifest <- data.frame(
  Motif_ID = redundant_ma_ids[matching_indices],
  TF_Name  = redundant_tf_names[matching_indices],
  stringsAsFactors = FALSE
)

# Remove any absolute duplicate rows if they exist
final_redundant_manifest <- unique(final_redundant_manifest)

# --- Print Summary Data ---
cat("Total motifs found in the REDUNDANT JASPAR file for your 681 TFs:", nrow(final_redundant_manifest), "\n")

# View the head of your new final table
head(final_redundant_manifest)
# Total is 733 motifs

```

TOTAL IS 733 MOTIFS

#### Look into

```{r}
h5_path <- "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/data/MarginalFootprints/Lactotrophs_FINAL_TOTAL_footprints.h5"
h5_structure <- h5ls(h5_path)
h5_ma_ids <- unique(h5_structure$name)
```

```{r}

motifs_missing_from_h5 <- setdiff(final_redundant_manifest$Motif_ID, h5_ma_ids)
missing_from_h5_df <- final_redundant_manifest[final_redundant_manifest$Motif_ID %in% motifs_missing_from_h5, ]

cat("--- CROSS-REFERENCE SUMMARY ---\n")
cat("Total expected motifs (from redundant list):", length(final_redundant_manifest$Motif_ID), "\n")
cat("Total unique keys found in .h5 file:", length(h5_ma_ids), "\n")
cat("Number of target motifs MISSING from .h5:", length(motifs_missing_from_h5), "\n")
cat("Number of unique TF genes impacted by these missing motifs:", length(unique(missing_from_h5_df$TF_Name)), "\n\n")

# Preview the missing TFs and their specific missing motif versions
cat("Preview of missing motifs and TFs:\n")
print(head(missing_from_h5_df, 15))

```

So many motifs are missing, and because different footprinting was conducted for the different cell types, this missing list will not be consistent across all cell types. Best practice would be to just rerun the footprinting as a new pipeline.

#### Make new tsvs for total list

```{r}

motif_sequences <- read.delim(
  "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/motif_sequences.tsv",
  header = FALSE,
  sep = "\t", 
  stringsAsFactors = FALSE
)
colnames(motif_sequences)[1] <- "Motif_ID"

final_filtered_sequences <- motif_sequences[motif_sequences$Motif_ID %in% final_redundant_manifest$Motif_ID, ]
cat("Total rows in original TSV:", nrow(motif_sequences), "\n")
cat("Rows matching final redundant list:", nrow(final_filtered_sequences), "\n\n")

# Split into 3 - Marginal Footprinting Crashes for the entire list of 733 
group_vector <- cut(seq_len(nrow(final_filtered_sequences)), breaks = 3, labels = FALSE)
split_lists <- split(final_filtered_sequences, group_vector)
list_1 <- split_lists[[1]]
list_2 <- split_lists[[2]]
list_3 <- split_lists[[3]]

write.table(
  list_1, 
  "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/PCA_motif_sequences_1.tsv", 
  sep = "\t", 
  row.names = FALSE, 
  col.names = FALSE,
  quote = FALSE
)

write.table(
  list_2, 
  "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/PCA_motif_sequences_2.tsv", 
  sep = "\t", 
  row.names = FALSE, 
  col.names = FALSE,
  quote = FALSE
)

write.table(
  list_3, 
  "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/PCA_motif_sequences_3.tsv", 
  sep = "\t", 
  row.names = FALSE, 
  col.names = FALSE,
  quote = FALSE
)


```

## Filter for TFs only present in stem cells

This section filters for TFs present in adult stem cells, for filtering purposes in the `marginal_footprint_analysis/3_MF_PCA_analysis.ipynb` file when assessing footprint differences between neonatal and adult stem cells.

```{r}

# Load files
RNA_exp_celltype <- read.csv("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/rawdata/median_RNAexpression_per_celltype.csv")
# Filter for TFs 
JASPAR_tf_motifs <- read_meme("/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/JASPAR_CORE_2026_non-redundant.meme")

# Assign rownames to gene ID
rownames(RNA_exp_celltype) <- RNA_exp_celltype$X
RNA_exp_celltype$X <- NULL

# See overlapping TF names
TF_alt_names <- sapply(JASPAR_tf_motifs, function(x) x@altname)
TF_alt_names_lowered <- tolower(TF_alt_names)
matching_rows <- rownames(RNA_exp_celltype)[tolower(rownames(RNA_exp_celltype)) %in% TF_alt_names_lowered]
missing_rows <- setdiff(TF_alt_names_lowered, tolower(rownames(RNA_exp_celltype)))
znf_rows <- grep("znf", missing_rows, value = TRUE)
combo_motif_rows <- grep("::", missing_rows, value = TRUE)
rest_of_missing_motifs <- setdiff(missing_rows, c(znf_rows, combo_motif_rows))

# Print results
cat("There are ", length(TF_alt_names_lowered), "in the JASPAR vertebrate non-redundant TF list.\nBut this includes multiple motifs per TF if there are multiple.\n")
cat("There are only really", length(unique(TF_alt_names_lowered)), "TFs in this list.\n")
cat("Found", length(matching_rows), "matching TFs.\n\n")
cat("There are", length(missing_rows), "missing TFs from the RNA-seq expression list.\n")
# Only 952 TFs really...
# Missing rows are zinc fingers, and dimer motifs. To be kept in downstream analysis. 
cat("There are", length(znf_rows), "zinc finger proteins within this list of missing TFs.\n")
cat("There are", length(combo_motif_rows), "motif combination TFs in this list of missing TFs.\n")
cat("The rest of the missing motifs are:", rest_of_missing_motifs, "\n")
cat("All missing TF motifs will still be incldued in the footprinting.")
```

```{r}

RNA_exp_celltype_TF <- RNA_exp_celltype[matching_rows, , drop = FALSE]

# Filter for RNA median expr >0.4 in stem cells
RNA_exp_celltype_filtered <- RNA_exp_celltype_TF[RNA_exp_celltype_TF$Stem_cells > 0.4, ]
head(RNA_exp_celltype_filtered) # 394 total
```

```{r}

# List of expressed TFs 
expressed_tfs_lowered <- tolower(rownames(RNA_exp_celltype_filtered))
# Missing TFs
missing_tfs_lowered <- unique(missing_rows)
# Combine for FINAL LIST
final_tfs_lowered <- c(expressed_tfs_lowered, missing_tfs_lowered)

print(final_tfs_lowered)

```

```{r}
# Save file
df_final_tfs <- data.frame(gene_symbol = final_tfs_lowered)
write.csv(df_final_tfs, 
          file = "/Users/georgiagoddard/Desktop/Research_Project/TF_mapping/metadata/stem_cells_adult_TF_expr_list.csv", 
          row.names = FALSE)

```
